### 2014-10-03 paper/include/background.tex
workn

To write to a row the corresponding wordline is first activated. As during a read, the transistors then connects the capacitors to the bitlines and charge will either flow from or to the capacitor. The row buffer are then loaded with the data that are driven to the bitlines by the sense amplifiers, thus storing the information on the capacitors. The row then remains open, as by a read, until another row shall be accessed, then the row is closed which disconnects the capacitors from the bitlines.

Over time the capacitors leaks charge and the capacitor has to be recharged in order to preserve the data. The operation to recharge a row is called refresh and is performed by opening a row to let the sense amplifiers amplify the detected voltage change and thus recharge the cells to either \(V_{DD}\) or 0 V. This follows the same procedure as a read operation and thus is the row refreshed when a read is performed.

The memory controller (MC) normally issues the refresh operation, the auto-refresh command, periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications. The interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C. The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced.

@@

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row. The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the 64ms interval. The second is called CAS before RAS refresh (CBR). In this scheme the memory module has an internal address counter that displays which row that shall be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. The advantage with this method is that the address is never put on the bus, which gives a lower power consumption and less bus usage than ROR. CBR is the most used method and is the one normally implemented in the auto-refresh command.

When using auto-refresh (CBR), all banks in a rank are blocked for normal accesses. All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank which is getting refreshed.

\begin{figure}[t]
    \centering
    \subfigure[DRAM throughput loss]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_throughput}
        \label{fig:dram_throughput}
    }
    \subfigure[DRAM power consumption]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_power}
        \label{fig:dram_power}
    }
    \caption{Projections on DRAM throughput loss and power consumption \cite{raidr}.}
    \label{fig:dram_data_proj}
\end{figure}

@@

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access. MENTION TIMING ISSUES IF ROWS GET LONGER.

When the DRAM chips become larger the throughput loss increases exponentially as \reffig{fig:dram_throughput} shows. Also when additional rows are added to the chip, more refresh commands has to be sent and of course gives this many more actual refreshes, especially if CBR is used. More refreshes means larger power consumption and is is also so that the refresh bocomes a growing portion of the total DRAM energy consumption. In \reffig{fig:dram_power} is is seen that when the device capacity reaches 64 GB the refresh power will nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

>>>>

To write to a row the corresponding wordline is first activated. As during a read, the transistors then connects the capacitors to the bitlines and charge will either flow from or to the capacitor. The row buffer are then loaded with the data that are driven to the bitlines by the sense amplifiers, thus storing the information on the capacitors. The row will be closed when another row shall be accessed.

As mentioned, the capacitors leaks charge over time and has to be recharged in order to preserve the data. The refresh is performed by opening a row to let the sense amplifiers amplify the detected voltage change and thus recharge the cells to either \(V_{DD}\) or 0 V. This follows the same procedure as a read operation and thus is the row refreshed when a read is performed.

The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications [source]. This interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C. %The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced. % Maybe not needed, we shall see.

@@

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row. The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the 64ms interval. The second is called CAS before RAS refresh (CBR). In this scheme the memory module has an internal address counter that displays which row that shall be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. The advantage with this method is that the address is never put on the bus, which gives a lower power consumption and less bus usage than ROR.

When using CBR, all banks in a rank are blocked for normal accesses. All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank which is getting refreshed.

\begin{figure}[t]
    \centering
    \subfigure[DRAM throughput loss]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_throughput}
        \label{fig:dram_throughput}
    }
    \subfigure[DRAM power consumption]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_power}
        \label{fig:dram_power}
    }
    \caption{Projections on DRAM throughput loss and power consumption in extended-temprature operation \cite{raidr}.}
    \label{fig:dram_data_proj}
\end{figure}

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.

As refresh operations are made by row, the refresh operations will increase liniary with the device capacity. In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}. Additional refresh operations gives larger power consumption and the refreshes becomes a growing portion of the total DRAM energy consumption as seen in \reffig{fig:dram_power}. When the device capacity reaches 64 GB the refresh power nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

@@

### 2014-10-03 paper/include/reducing_refresh_power.tex
workn

\subsection{Approaches to reduce refresh power}
\todo[inline]{Replace the copy-paste texts}

\subsubsection*{\textbf{Access Recency}}
DRAM accesses imply refresh operations, and so a subsequent refresh to the same row can be postponed. The number of rows affected depends on how many different rows are accessed within the maximum refresh period, which may be few for many workloads. Ghosh et al. [9] propose Smart Refresh, which divides the refresh period into phases and maintains a per-row timeout counter in the MC. The MC decrements the counters each phase and issues a RAS-only refresh when a counter hits zero.

\subsubsection*{\textbf{Retention Time}}
Retention time refers to the period during which cells hold valid values as charge gradually leaks. Process variation [10][16] causes the retention time of DRAM cells to vary across the chip. Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.

\subsubsection*{\textbf{Data Tolerance}}
Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results. Approximate computation [31, 8] exploits this approximate data to realize tradeoffs among performance, energy, and accuracy. Cells containing such error-tolerant T data need not be refreshed as often as those containing critical data. How many cells fall into this category depends on application characteristics.

\subsubsection*{\textbf{Validity}}
If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful. Severa
l software approaches thus attempt to trigger refreshes only for rows with valid data. The effectiveness of schemes using such V refresh information is sensitive to the total memory usage of the system.

>>>>

\subsection{Approaches to reduce refresh power}

\subsubsection*{\textbf{Access Recency}}
As mentioned, a DRAM access is performed in the same way as a refresh, which makes it possible to delay the next refresh operation to an accessed row without endangering the stored data. All accessed rows within the maximum refresh period result in that a refresh can be postponed. This implies that if the system seldom accesses the DRAM data, or only a small portion of it, the approach's gain diminishes.

\subsubsection*{\textbf{Retention Time}}
.

\todo[inline]{Replace the copy-paste texts}
Retention time refers to the period during which cells hold valid values as charge gradually leaks. Process variation [10][16] causes the retention time of DRAM cells to vary across the chip. Note that approaches that ex-ploit R information are insensitive to system workloads and global memory usage, which makes them attractive omponents for a refresh-optimized memory subsystem.

\subsubsection*{\textbf{Data Tolerance}}
.

\todo[inline]{Replace the copy-paste texts}
Applications like games, media processing, machine learning, and unstructured information analysis tolerate errors in portions of their data and still produce acceptably accurate results. Approximate computation [31, 8] exploits this approximate data to realize tradeoffs among performance, energy, and accuracy. Cells containing such error-tolerant T data need not be refreshed as often as those containing critical data. How many cells fall into this category depends on application characteristics.

\subsubsection*{\textbf{Validity}}
.

\todo[inline]{Replace the copy-paste texts}
If the OS has not allocated the page frame, its data are meaningless, and refreshes to it are wasteful. Severa
l software approaches thus attempt to trigger refreshes only for rows with valid data. The effectiveness of schemes using such V refresh information is sensitive to the total memory usage of the system.
